Automating Free Logic in HOL, with an Experimental Application in Category Theory

### 2017-04-29 FreeLogic/2017-Journal/article.tex
update

%\titlerunning{Short form of title}        % if too long for running head

@@

In the first part of this article (\S2 and \S3) we show how free logic
can be elegantly ``implemented'' in any theorem proving system for
classical higher-order logic (HOL) \cite{B5}. The proposed solution
employs a semantic embedding of free in HOL. We
present an exemplary implementation of this idea in the mathematical
proof assistant Isabelle/HOL \cite{NPW02}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool \cite{Sledgehammer}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners. As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

@@

@@

The stepwise evolution has been described in detail in\cite{ArXiv}. Below we
summarize these experiments.  However, first we describe some basic
modeling decisions for the technical encoding in Isabelle/HOL. The
sources of our experiments, as discussed in \cite{ArXiv}, are available at
\url{http://christoph-benzmueller.de/papers/2017-arXiv.zip}. Note that in these
sources we did add neither the definite description operator nor the
designated undefined object $\hol{\star}$, since both were not required in
this particular application.

@@

This definition was suggested by an exercise in the textbook by Freyd
and Scedrov \cite{FreydScedrov90}
on p.~4.  In earlier experiments we used a longer definition which can
be proved equivalent on the basis of the other axioms. For monoids,
where composition is total, $\id i$ means $i$ is a two-sided identity
— and such are unique. For categories the property is much weaker.

@@

\subsection{Remark on the Experiments}
All proofs above and all proofs in the rest of this paper (unless
stated otherwise) have been obtained fully automatically in very
reasonable time (typically just a few seconds) with the
Sledgehammer tool in Isabelle/HOL. This tool interfaces to prominent
first-order automated theorem provers such as CVC4 \cite{CVC4}, Z3
\cite{Z3}, E \cite{E} and Spass \cite{Spass}.  Remotely, also
  provers such as Vampire \cite{Vampire}, or the higher-order provers
  Satallax \cite{Satallax} and LEO-II \cite{LEO} can be reached. For
  example, to prove axiom $E_{iii}$ from Axioms Set II, we have called
  Sledgehammer on all axioms of Axioms Set II. The provers then, via
  Sledgehammer, suggested to call trusted/verified tools in
  Isabelle/HOL with the exactly required dependencies they
  detected. With the provided dependency information the trusted tools
  in Isabelle/HOL were then able to reconstruct the external proofs on
  their own.  This way we obtain a verification of our claims in
  Isabelle/HOL, in which all the proofs have nevertheless been
  contributed by automated theorem provers. However, there are still
  a few calls of the \textsc{smt}-tactic contained in our document, 
   which we are not entirely trivial to replace. In this sense, our
   Isabelle/HOL verification is modulo the correctness of the \textsc{smt} solvers
   CVC4 \cite{CVC4} and Z3 \cite{Z3}.

@@

\subsection{Missing Strictness Axioms in Alternative
  Setting} \label{subsec-freyd-scedrov-2} We study the axiom system by
Freyd and Scedrov once again. However, this time we restrict the free
variables in their system to range over existing objects only. In the
context of algebraic theories, it could be argued that this is the
preferred reading of free variables.  By employing the free logic
universal quantifier $\forall$, which realizes such a restriction, 
we thus modify Axioms Set FS-II into
Axioms-Set FS-III as displayed in Table~\ref{axioms-sets-3}.

For Axioms Set FS-III the consistency checks with Nitpick succeeds,
even if we assume undefined objects. However, this axioms set is
obviously weaker than Axioms Set V from Table
\ref{axioms-sets-1}. In fact, as has been shown by Nitpick, none of
the axioms of this set are implied.  The situation changes when we
explicitly postulate strictness of $\dom$, $\cod$ and $\comp$. Doing
so we obtain Axioms Set FS-IV from Table \ref{axioms-sets-3}, which,
as Nitpick confirms, is consistent even if we assume undefined
objects. And the automated theorem provers via Sledgehammer confirm
that Axioms Set FS-IV is equivalent to Axioms Set V, as
intended. Unfortunately, however, respective strictness conditions are
not mentioned in the textbook by Freyd and Scedrov.

\begin{table} \centering \normalsize
\begin{tabular}{ll}
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-III: Freyd and Scedrov in our notation (with issues)} \\
%\hline
\\
  $B1$  & $\all{x}\all{y} \ex (x \comp y) \longleftrightarrow \dom x \cong \cod y$ \\
  $B2a$ & $\all{x} \cod(\dom x) \cong \dom x$ \\  
  $B2b$ & $\all{y} \dom(\cod y) \cong \cod y$ \\  
  $B3a$  & $\all{x} x \comp (\dom x) \cong x$ \\ 
  $B3b$ & $\all{y} (\cod y) \comp y \cong y$ \\
  $B4a$ & $\all{x}\all{y} \dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $B4b$ & $\all{x}\all{y} \cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $B5$   & $\all{x}\all{y}\all{z} x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-IV: Freyd and Scedrov in our
  notation (without issues)} \\
%\hline
\\
 $B0a$ & $\ex (x \comp y) \longrightarrow  (\ex x \wedge \ex y)$ \\
 $B0b$ & $\ex (\dom x)  \longrightarrow   \ex x$ \\
 $B0b$ & $\ex (\cod x)  \longrightarrow   \ex x$ \\
  $B1$  & $\all{x}\all{y} \ex (x \comp y) \longleftrightarrow \dom x \cong \cod y$ \\
  $B2a$ & $\all{x} \cod(\dom x) \cong \dom x$ \\  
  $B2b$ & $\all{y} \dom(\cod y) \cong \cod y$ \\  
  $B3a$  & $\all{x} x \comp (\dom x) \cong x$ \\ 
  $B3b$ & $\all{y} (\cod y) \comp y \cong y$ \\
  $B4a$ & $\all{x}\all{y} \dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $B4b$ & $\all{x}\all{y} \cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $B5$   & $\all{x}\all{y}\all{z} x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\end{tabular}
\caption{The axioms set of Freyd and Scedrov in our
  notation and with variable restriction to existing objects only.\label{axioms-sets-3}}
\end{table}

\section{Summary and Further Work}
We have developed a new reasoning framework for free logic, and we
have experimentally applied it for some first experiments in category
theory.  We have demonstrated how modern proof assistants and theorem
provers for classical higher-order logic may well support the
reasoning in free logic. More concretely, we have applied our new free
logic reasoning framework for the systematic exploration of axiom
systems for category theory. Without tools support such experiments
would be extremely tedious and error prone.
In the course of our experiments, automated theorem provers 
have revealed some (minor) issue in the textbook of Freyd and Scedrov \cite{FreydScedrov90},
which we were able to correct. The correction essentially corresponds
to the axiom system by Scott proposed earlier \cite{Scott67}.
All our findings were achieved directly by or in
close interaction with automated reasoning tools.

@@

>>>>

\titlerunning{Automating Free Logic in HOL}        % if too long for running head

@@

In the first part of this article (\S2 and \S3) we show how free logic
can be elegantly ``implemented'' in any theorem proving system for
classical higher-order logic (HOL) \cite{B5}. The proposed solution
employs a semantic embedding of free in HOL. We present, as an
example, one implementation of this idea in the mathematical proof
assistant Isabelle/HOL \cite{NPW02}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool \cite{Sledgehammer}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners. As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

@@

The above results enable the employment of any theorem prover that
supports $\HOL$ with definite description to reason with $\FFOL$,
including TPTP THF \cite{J22} compliant systems such as Satallax,
Nitpick, LEO-II and Leo-III. Alternatively, this theory can be encoded
in interactive proof assistants, such as Isabelle/HOL. <
  % \footnote{An example of respective abstract level proof
  % tactics in the context of a semantical embedding of an
  % hyperintensional higher-order modal logic in $\HOL$ can be found at:
  % \url{https://github.com/ekpyron/TAO}; in this document a
  % computer-formalization of Ed Zalta's Principia
  % Logico-Metaphysica \url{https://mally.stanford.edu/principia.pdf} is presented.}

@@

The stepwise evolution has been described in detail
in\cite{ArXiv}. Below we summarize these experiments.  However, first
we describe some basic modeling decisions for the technical encoding
in Isabelle/HOL. The sources of our experiments, as discussed in
\cite{ArXiv}, are available at
\url{http://christoph-benzmueller.de/papers/2017-arXiv.zip}. Note that
in these sources we added neither the definite description operator
nor the designated undefined object $\hol{\star}$, since both were not
required in this particular application.

@@

This definition was suggested by an exercise in the textbook by Freyd
and Scedrov \cite{FreydScedrov90}
on p.~4.  In earlier experiments we used a longer definition which can
be proved equivalent on the basis of the other axioms. For monoids,
where composition is total, $\id i$ means $i$ is a two-sided identity
--- and such are unique. For categories the property is much weaker.

@@

\subsection{Remark on the Experiments}
All proofs above and all proofs in the rest of this paper (unless
stated otherwise) have been obtained fully automatically in very
reasonable time (typically just a few seconds) with the
Sledgehammer tool in Isabelle/HOL. This tool interfaces to prominent
first-order automated theorem provers such as CVC4 \cite{CVC4}, Z3
\cite{Z3}, E \cite{E} and Spass \cite{Spass}.  Remotely, also
  provers such as Vampire \cite{Vampire}, or the higher-order provers
  Satallax \cite{Satallax} and LEO-II \cite{LEO} can be reached. For
  example, to prove axiom $E_{iii}$ from Axioms Set II, we have called
  Sledgehammer on all axioms of Axioms Set II. The provers then, via
  Sledgehammer, suggested to call trusted/verified tools in
  Isabelle/HOL with the exactly required dependencies they
  detected. With the provided dependency information the trusted tools
  in Isabelle/HOL were then able to reconstruct the external proofs on
  their own.  This way we obtain a verification of our claims in
  Isabelle/HOL, in which all the proofs have nevertheless been
  contributed by automated theorem provers. However, there are still
  a few calls of the \textsc{smt}-tactic contained in our document, 
   which we found are not entirely trivial to replace. In this sense, our
   Isabelle/HOL verification is modulo the correctness of the \textsc{smt} solvers
   CVC4 \cite{CVC4} and Z3 \cite{Z3}.

@@

\begin{table} \centering \normalsize
\begin{tabular}{ll}
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-III: Freyd and Scedrov in our notation (with issues)} \\
%\hline
\\
  $B1$  & $\all{x}\all{y} \ex (x \comp y) \longleftrightarrow \dom x \cong \cod y$ \\
  $B2a$ & $\all{x} \cod(\dom x) \cong \dom x$ \\  
  $B2b$ & $\all{y} \dom(\cod y) \cong \cod y$ \\  
  $B3a$  & $\all{x} x \comp (\dom x) \cong x$ \\ 
  $B3b$ & $\all{y} (\cod y) \comp y \cong y$ \\
  $B4a$ & $\all{x}\all{y} \dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $B4b$ & $\all{x}\all{y} \cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $B5$   & $\all{x}\all{y}\all{z} x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-IV: Freyd and Scedrov in our
  notation (without issues)} \\
%\hline
\\
 $B0a$ & $\ex (x \comp y) \longrightarrow  (\ex x \wedge \ex y)$ \\
 $B0b$ & $\ex (\dom x)  \longrightarrow   \ex x$ \\
 $B0b$ & $\ex (\cod x)  \longrightarrow   \ex x$ \\
  $B1$  & $\all{x}\all{y} \ex (x \comp y) \longleftrightarrow \dom x \cong \cod y$ \\
  $B2a$ & $\all{x} \cod(\dom x) \cong \dom x$ \\  
  $B2b$ & $\all{y} \dom(\cod y) \cong \cod y$ \\  
  $B3a$  & $\all{x} x \comp (\dom x) \cong x$ \\ 
  $B3b$ & $\all{y} (\cod y) \comp y \cong y$ \\
  $B4a$ & $\all{x}\all{y} \dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $B4b$ & $\all{x}\all{y} \cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $B5$   & $\all{x}\all{y}\all{z} x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\end{tabular}
\caption{The axioms set of Freyd and Scedrov in our
  notation and with variable restriction to existing objects only.\label{axioms-sets-4}}
\end{table}

\subsection{Missing Strictness Axioms in Alternative
  Setting} \label{subsec-freyd-scedrov-2} We study the axiom system by
Freyd and Scedrov once again. However, this time we restrict the free
variables in their system to range over existing objects only. In the
context of algebraic theories, it could be argued that this is the
preferred reading of free variables.  By employing the free logic
universal quantifier $\forall$, which realizes such a restriction, 
we thus modify Axioms Set FS-II into
Axioms-Set FS-III as displayed in Table~\ref{axioms-sets-4}.

For Axioms Set FS-III the consistency checks with Nitpick succeeds,
even if we assume undefined objects. However, this axioms set is
obviously weaker than Axioms Set V from Table
\ref{axioms-sets-1}. In fact, as has been shown by Nitpick, none of
the axioms of this set are implied.  The situation changes when we
explicitly postulate strictness of $\dom$, $\cod$ and $\comp$. Doing
so we obtain Axioms Set FS-IV from Table \ref{axioms-sets-4}, which,
as Nitpick confirms, is consistent even if we assume undefined
objects. And the automated theorem provers via Sledgehammer confirm
that Axioms Set FS-IV is equivalent to Axioms Set V, as
intended. Unfortunately, however, respective strictness conditions are
not mentioned in the textbook by Freyd and Scedrov.

\section{Summary and Further Work}
We have developed a new reasoning framework for free logic, and we
have experimentally applied it for some first experiments in category
theory.  We have demonstrated how modern proof assistants and theorem
provers for classical higher-order logic may well support the
reasoning in free logic. More concretely, we have applied our new free
logic reasoning framework for the systematic exploration of axiom
systems for category theory. Without tools, support of such experiments
would be extremely tedious and error prone.
In the course of our experiments, automated theorem provers 
have revealed some (minor) issue in the textbook of Freyd and Scedrov \cite{FreydScedrov90},
which we were able to correct. The correction essentially corresponds
to the axiom system by Scott proposed earlier \cite{Scott67}.
All our findings were achieved directly by or in
close interaction with automated reasoning tools.

@@

Moreover, abstract level proof tactics could be provided in proof
assistants to support --- as an alternative to always unfolding the
mapping from $\FFOL$ to $\HOL$ --- intuitive interaction (and even
automation) in $\FFOL$ on top the semantical embedding.
